---
title: "Analysis workflow with fatTissueTumor"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Analysis workflow with fatTissueTumor}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
library(fatTissueTumor)
```

# Introduction

This vignette shows a complete analysis workflow with the **fatTissueTumor** package, from data import to preprocessing, co-expression analysis, enrichment and visualization.


# Main functions

|Funzione               | Short Description                                         |
|-----------------------|-----------------------------------------------------------|
|import_GEO             | Import data from GEO                                      |
|import_TCGA            | Import data from TCGA                                     |
|import_ArrayExpress    |	Import data from ArrayExpress                             |
|preprocess_expression  |	Advanced preprocessing pipeline                           |
|gene_coexpression	    | Calculate the correlation between two genes               |
|top_gene_coexpression  |	Identifies the most co-expressed gene pairs               |
|plot_enrichment_result |	View enrichment results (GO/KEGG/Reactome)                |

# Importing sample data
```{r load_data, eval=FALSE}
data("expr_breast")
data("pheno_breast")
str(expr_breast)
str(pheno_breast)
```

# Preprocessing
```{r preprocessing, eval=FALSE}
res <- preprocess_expression(expr_breast, log_transform = TRUE, return_report = TRUE)
expr_proc <- res$expr
print(res$report)
```

# Co-expression analysis
```{r co_expression, eval=FALSE}
# Calculate the correlation between two genes.
gene_coexpression(expr_proc, "LEP", "STAT3")

# Find the 5 most co-expressed pairs.
top_pairs <- top_gene_coexpression(expr_proc, top_n = 5)
print(top_pairs)
```

# GO enrichment analysis
```{r enrichment, eval=FALSE}
if (!requireNamespace("clusterProfiler", quietly = TRUE)) BiocManager::install("clusterProfiler")
if (!requireNamespace("org.Hs.eg.db", quietly = TRUE)) BiocManager::install("org.Hs.eg.db")
library(clusterProfiler)
library(org.Hs.eg.db)

# Select the most highly expressed genes as an example.
gene_symbols <- rownames(expr_proc)[1:10]
gene_df <- bitr(gene_symbols, fromType = "SYMBOL", toType = "ENTREZID", OrgDb = org.Hs.eg.db)
gene_list <- unique(gene_df$ENTREZID)

ego <- enrichGO(
  gene = gene_list,
  OrgDb = org.Hs.eg.db,
  keyType = "ENTREZID",
  ont = "BP",
  pAdjustMethod = "BH",
  pvalueCutoff = 0.05,
  qvalueCutoff = 0.2
)

# Visualization of enrichment results
plot_enrichment_result(ego, type = "dotplot", showCategory = 10)
```

# Conclusions
The **fatTissueTumor** package enables complete workflows for the analysis of adipose tissue-tumor interaction, from data collection to visualization of results.

For more details, see the documentation of individual functions (?function_name) or the README of the package.
