---
title: "Pipeline"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Pipeline}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(fatTissueTumor)
library(clusterProfiler)
```

## Loading sample data
```{r load_data, include=TRUE}
# Load the sample dataset included in the package.
data("expr_small")
dim(expr_small)
head(expr_small[, 1:5])
```

## Normalization
```{r normalization, include=TRUE}
# Normalization z-score
expr_z <- normalize_expression(expr_small, method = "zscore")
summary(as.vector(expr_z))
```


## Filtering Genes
```{r filtering, include=TRUE}
# Filters low-expression/variance genes
expr_filt <- filter_genes(expr_z, min_mean = 0, min_var = 0.5)
dim(expr_filt)
```


## Boxplot before and after normalization
```{r Boxplot, include=TRUE,fig.width=7, fig.height=5}
par(mfrow = c(1, 2))
boxplot(expr_small, main = "Original", las = 2, col = "lightblue")
boxplot(expr_z, main = "Z-score", las = 2, col = "lightgreen")
par(mfrow = c(1, 1))
```


## Density of expressions
```{r density, include=TRUE,fig.width=7, fig.height=5}
plot(density(as.vector(expr_small)), main = "Density values expression (original)", col = "blue")
lines(density(as.vector(expr_z)), col = "green")
legend("topright", legend = c("Original", "Z-score"), col = c("blue", "green"), lwd = 2)
```


## PCA 
```{r PCA, include=TRUE,fig.width=7, fig.height=5}
pca_orig <- prcomp(t(expr_small), scale. = FALSE)
pca_z <- prcomp(t(expr_z), scale. = FALSE)

par(mfrow = c(1, 2))
plot(pca_orig$x[,1:2], main = "PCA Original", col = "blue", pch = 19, xlab = "PC1", ylab = "PC2")
plot(pca_z$x[,1:2], main = "PCA Z-score", col = "green", pch = 19, xlab = "PC1", ylab = "PC2")
par(mfrow = c(1, 1))
```


## Heatmap of the most variable genes.
```{r Heatmap, include=TRUE,fig.width=7, fig.height=5}
# Select the 20 genes with highest variance.
gene_vars <- apply(expr_z, 1, var)
top_genes <- names(sort(gene_vars, decreasing = TRUE))[1:20]
heatmap_data <- expr_z[top_genes, ]

heatmap(as.matrix(heatmap_data), scale = "row", 
        main = "Heatmap: 20 genes plus variables (z-score)",
        col = colorRampPalette(c("navy", "white", "firebrick3"))(50))
```


## Analysis of the most co-expressed gene pairs.
```{r Pairs_Coexpr, include=TRUE,fig.width=7, fig.height=5}
coexp <- top_gene_coexpression(expr_filt, method = "pearson", top_n = 5)
coexp
```


## Visualization of the first pairs of most co-expressed genes Positive
```{r plot_coexpr_pairs_pos, fig.width=7, fig.height=4}
# View the top 3 scatterplots of the most positively co-expressed pairs.
plot_top_gene_pairs(expr_filt, coexp, n_plot = 3, point_col = "purple", point_cex = 2)
```


## Visualization of the first most co-expressed gene pairs Negative
```{r plot_coexpr_pairs_neg, fig.width=7, fig.height=4}
neg_coexp <- coexp[order(coexp$correlation), ][1:3, ]
plot_top_gene_pairs(expr_filt, neg_coexp, n_plot = 3, point_col = "orange", point_cex = 2)
```


## Automatic saving of graphs
```{r saving, eval=FALSE}
# This line saves the graphs as PNG files in the workbook
plot_top_gene_pairs(expr_filt, coexp, n_plot = 3, save_plot = TRUE, prefix = "coexp_pos_")
plot_top_gene_pairs(expr_filt, neg_coexp, n_plot = 3, save_plot = TRUE, prefix = "coexp_neg_")
*(Nota: `eval=FALSE` prevents the chunk from being run during the compilation of the vignette, but you can run it manually in console.)*
```


## Plotly
```{r plotly, eval=FALSE, fig.width=7, fig.height=4}
# Run only if you have plotly installed
plot_gene_pair_interactive(expr_filt, coexp$gene1[1], coexp$gene2[1])
```


## Automatic GO enrichment analysis on probe IDs.
```{r enrich_automatic, fig.width=8, fig.height=5}
gene_list <- rownames(expr_filt)[1:40]
go_res <- enrichment_auto_hgu95(gene_list, type = "GO")
plot_enrichment(go_res, top = 10)
```


## Interactive heatmap of the most variable genes.
```{r heatmaply, eval=TRUE, message=FALSE, warning=FALSE, fig.width=7, fig.height=4}
interactive_heatmap(expr_small, top_genes = 30)
```


## Gene co-expression network
```{r coexp-network, eval=TRUE, message=FALSE, warning=FALSE, fig.width=7, fig.height=4}
g <- coexpression_network(expr_small, top_genes = 30, method = "pearson", threshold = 0.7)
plot_coexpression_network(g)
plot_coexpression_network_interactive(g)
```


## Advanced interactive heatmap
```{r heatmaply-advanced, eval=TRUE, message=FALSE, warning=FALSE, fig.width=7, fig.height=4}
group <- rep(c("Lean", "Obese"), length.out = ncol(expr_small))
sample_anno <- data.frame(Group = group)
rownames(sample_anno) <- colnames(expr_small)

interactive_heatmap(expr_small,
                   top_genes = 30,
                   dist_method = "manhattan",
                   hclust_method = "average",
                   colors = heatmaply::cool_warm,
                   sample_annotation = sample_anno)
```


## Custom co-expression network
```{r coexp-network-custom, eval=TRUE, message=FALSE, warning=FALSE, fig.width=7, fig.height=5}
group <- rep(c("Lean", "Obese"), length.out = nrow(expr_small))
names(group) <- rownames(expr_small)
g <- coexpression_network(expr_small, top_genes = 30, method = "pearson", threshold = 0.7)
plot_coexpression_network(g, node_anno = group, palette = heat.colors)
```


## GO enrichment analysis and visualization.
```{r enrichGO-setup, message=FALSE, warning=FALSE, fig.width=7, fig.height=5}
# Install packages only if they are missing (optional, you can remove these lines if already installed)
if (!requireNamespace("clusterProfiler", quietly = TRUE)) BiocManager::install("clusterProfiler")
if (!requireNamespace("org.Hs.eg.db", quietly = TRUE)) BiocManager::install("org.Hs.eg.db")
if (!requireNamespace("enrichplot", quietly = TRUE)) BiocManager::install("enrichplot")

library(clusterProfiler)
library(org.Hs.eg.db)
library(enrichplot)
# Example of gene symbols
gene_symbols <- c("TP53", "STAT3", "LEP", "JUN", "FOS", "EGFR", "AKT1", "MAPK1", "MAPK3", "PIK3CA")
# Convert Symbols -> EntrezID
gene_df <- bitr(gene_symbols, fromType = "SYMBOL", toType = "ENTREZID", OrgDb = org.Hs.eg.db)
gene_list <- unique(gene_df$ENTREZID)
# GO enrichment analysis
ego <- enrichGO(
  gene = gene_list,
  OrgDb = org.Hs.eg.db,
  keyType = "ENTREZID",
  ont = "BP",
  pAdjustMethod = "BH",
  pvalueCutoff = 0.05,
  qvalueCutoff = 0.2
)
# Visualization: dotplot
plot_enrichment_result(ego, type = "dotplot", showCategory = 10)
# Visualization: barplot
plot_enrichment_result(ego, type = "barplot", showCategory = 10)
# Visualization: cnetplot (network)
plot_enrichment_result(ego, type = "cnetplot", showCategory = 5)
```


## Advanced preprocessing pipeline
```{r preprocess-pipeline, eval=TRUE, message=FALSE, warning=FALSE}
batch <- rep(c("A", "B"), length.out = ncol(expr_small))
res <- preprocess_expression(
  expr_small,
  min_expr = 1,
  min_expr_frac = 0.2,
  min_var = 0.01,
  max_na_frac = 0.2,
  min_expr_sample = 1,
  min_var_sample = 0,
  max_na_frac_sample = 0.2,
  normalize = "zscore",
  log_transform = TRUE,
  batch = batch,
  return_report = TRUE
)
expr_proc <- res$expr
print(res$report)
```


## Importing public data
```{r import-geo, eval=FALSE, include=FALSE}
## Advanced import of public data
gse <- import_GEO("GSE2508", sample_type = "obese", gene_list = c("LEP", "STAT3"), normalize = TRUE, convert_id = "symbol")
str(gse)
expr_tcga <- import_TCGA("TCGA-BRCA")
dim(expr_tcga)
```
#' @examples
#' \dontrun{
#' # This example requires an internet connection and may fail during automated checks
#' ae <- import_ArrayExpress("E-MTAB-62", normalize = TRUE)
#' str(ae)
#' }
